<!--
%\VignetteEngine{knitr::knitr}
%\VignetteIndexEntry{Worst case time complexity}
-->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The time complexity of the algorithm depends on the number of
intervals (candidate changepoints stored). Here we compute the
mean number of intervals for real Mono27ac data, and synthetic
count data which are always increasing.

```{r}
data(Mono27ac, package="PeakSegDisk", envir=environment())
library(data.table)
loss.list <- list()
type.colors <- c(
  "Theoretical\nworst case"="grey",
  "Increasing"="red",
  "Mono27ac"="black")
dfac <- function(x){
  factor(x, names(type.colors))
}
N.data.vec <- 10^seq(1, 3)
for(penalty in c(0, 1e2, 1e4, 1e6)){
  for(N.data in N.data.vec){
    some.cov <- Mono27ac$coverage[1:N.data]
    some.inc <- data.table(some.cov)
    some.inc[, count := 1:.N]
    data.list <- list(
      Mono27ac=some.cov,
      Increasing=some.inc)
    for(data.type in names(data.list)){
      df <- data.list[[data.type]]
      fit <- PeakSegDisk::PeakSegFPOP_df(df, penalty)
      loss.list[[paste(penalty, N.data, data.type)]] <- data.table(
        N.data,
        data.type=dfac(data.type),
        fit$loss)
    }
  }
}
(loss <- do.call(rbind, loss.list))
```

Theoretically the most intervals that could be stored is $O(i)$ for
each data point $i\in\{1, ..., N\}$. Therefore the largest total
number of intervals is `sum(1:N)`, which can also be computed by
`N*(N+1)/2`. The largest mean is `mean(1:N)`, which can be computed
via `sum(1:N)/N` = `(N+1)/2`.
  
```{r}
(worst.dt <- data.table(
  N.data=N.data.vec,
  mean.intervals=(N.data.vec+1)/2,
  data.type=dfac("Theoretical\nworst case")))
```

The plot below shows that the algorithm achieves the theoretical
worst case time complexity for the synthetic increasing data,
when the penalty is large. But the number of intervals is always
much smaller for real Mono27ac ChIP-seq data.
  
```{r}
ggplot()+
  theme_bw()+
  theme(panel.margin=grid::unit(0, "lines"))+
  facet_grid(. ~ penalty, labeller=label_both)+
  geom_line(aes(
    N.data, mean.intervals, color=data.type),
    data=worst.dt)+
  scale_color_manual(
    values=type.colors,
    breaks=names(type.colors))+
  geom_line(aes(
    bedGraph.lines, mean.intervals, color=data.type),
    data=loss)+
  scale_x_log10()+
  scale_y_log10()
```
